---
title: "Product Insight Sales"
output: html_notebook
---

```{r}
#rm(list = ls())
```

## libaries used for the project 

```{r error=TRUE}
library(DBI)
library (RSQLite)
library(tidyverse)
library(tibble)
library(ggplot2)
library(rpart)
library(rpart.plot)
library(caret)
library(corrplot)
```


# Connecting to the Database

The dataset is stored in a SQLite database file named **`project1_raw_data.db`**, which is saved in the Desktop folder. The `RSQLite` package is used to establish a connection to the database. Once connected, the `PRAGMA table_info(weather)` query is executed to inspect the structure of the `weather` table, including its column names and data types.

```{r error=TRUE}
con <- dbConnect(
  RSQLite:: SQLite(),
  "/Users/AshishLama/Desktop/project1_raw_data.db"
)

data = dbGetQuery(con, "PRAGMA table_info(weather)") 
```


# Retrieving and Joining the Data

The following SQL query extracts the required variables by joining the `train`, `key`, and `weather` tables. The `train` table is linked to the `key` table using `store_nbr`, while the `weather` table is joined using both `station_nbr` and `date`. This creates a single dataset containing sales information together with the corresponding weather conditions for each store and date.

```{r error=TRUE}
df <- dbGetQuery(con, "

                 SELECT
                 t.date,
                 t.units,
                 t.store_nbr,
                 t.item_nbr,
                 w.station_nbr,
                 w.tavg,
                 w.preciptotal,
                 w.codesum,
                 w.tmax,
                 w.tmin,
                 w.dewpoint,
                 w.wetbulb,
                 w.heat,
                 w.cool,
                 w.sealevel,
                 w.resultdir
                 
                FROM train t
                
                JOIN key k ON t.store_nbr = k.store_nbr
                
                JOIN weather w ON k.station_nbr = w.station_nbr AND t.date = w.date
                 
                 ")

```


checking the dataframe

```{r}
df
# view(df)
```

# Closing the Database Connection

After the required data has been extracted, the database connection is closed to release system resources and ensure that the connection is properly terminated.

```{r}
dbDisconnect(con)
```


# Exploring the dataset

A summary of the dataset is generated to provide descriptive statistics and identify the distribution of each variable

```{r}
summary(df)
```


```{r}
sapply(df, class)
```
## DATA WANGLING

```{r}
df_clean <- df
```


```{r}
# creating the columns names under 'var'.
var <- c("tavg", "preciptotal", "tmax", "tmin", "dewpoint", "wetbulb", "heat", "cool", "sealevel", "resultdir")

```

we can see there are a lot of the NA's in the provided columns on var. we see if the there are NA and see the whitespacesand convert it into NA on the columns.
After that we replace the NA into the mean value.  

```{r}
mean_val <- colMeans(df_clean[, var], na.rm = TRUE)

#replacing NA with mean value of each column

for (i in var){
  # treating whitespace as NA
  df_clean[,i][df_clean[,i] == ""| df_clean[, i] == " "] <- NA
  
  #replacing the NA with mean
  df_clean[,i][is.na(df_clean[,i])] <- mean_val[i]
}
```

We can see that there are no NA in the provided columns.
```{r}
summary(df_clean)
```

checking the class of the columns.

```{r error=TRUE}
sapply(df_clean,class)
```

we can see the date is on the character we are going to change the class into Date.

```{r error=TRUE}

Sys.Date()
df_clean$date <- as.Date(df_clean$date, format = '%Y-%m-%d')

```


checking the class of the columns.

```{r error=TRUE}
sapply(df_clean,class)
```


From date we will be creating two features to do analysis.

```{r error=TRUE}

# creating the day features from the date.
df_clean$day <- weekdays(df_clean$date)

# creating the month features from the date
df_clean$month <- as.integer((format(df_clean$date, "%m")))

```


creating the weekends and weekdays, weekdays has monday,tuesday,wednesday,thrusday,friday and weekend has saturday and sunday.

```{r error=TRUE}

df_clean <- df_clean %>%
  mutate(
    is_weekends = as.integer((day %in% (c("Saturday", "Sunday")))),
    is_weekdays = as.integer(day %in% (c("Monday", "Tuesday","Wednesaday","Thrusday","Friday")))
  )

```


Black entries in the codesum variables are replaced with the 'moderate'.

```{r}
df_clean <- df_clean %>% 
  mutate(codesum = ifelse(codesum == "" | codesum == " ", "Moderate", codesum))
```

vewing the dataframe.

```{r}
df_clean
```

Creating Weather Indicator Variables

The 'codesum' variable is transform into binary indicators for the several weather conditions.
if the weather condition is presennt it will contain 1 else 0.

```{r}
df_clean <- df_clean %>%
  mutate(
    is_clear  = as.integer(grepl("Moderate", codesum)),
    is_rainy  = as.integer(grepl("RA|DZ|PY|SH", codesum)),
    is_snowy  = as.integer(grepl("SN|SG|GS|FZ|IC|PL", codesum)),
    is_foggy  = as.integer(grepl("FG|BR|HZ", codesum)),
    is_stormy = as.integer(grepl("DS|SS|BL|PO|DU", codesum))
  )

```

viewing the dataframe
```{r}
df_clean
```





## Filtering Data for the Top Three Items

The datasets is filtered to retain only observations corresponding to the three
highest-selling items. This subset is used for more focused analysis.

```{r error=TRUE}

top3_items <- c(5, 9, 45)

df_top3 <- df_clean %>%
  filter(item_nbr %in% top3_items)

```


viewing the df_top3 dataframe

```{r}
df_top3
```

## Visualisation

quantriles shows the average temperature.

```{r}
tavg_iqr <- quantile(df_top3$tavg)
tavg_iqr
```


box plot shows the Temperature Distribution(tavg)
```{r error=TRUE}
boxplot(df_top3$tavg,
        main = "Temperature Distribution (Tavg)",
        ylab = "Average Temperature",
        col = "Skyblue")
```

Average Temperature vs the frequency.
```{r error=TRUE}
hist(df_top3$tavg,
     main = "Histogram of Average Temperature",
     xlab = "Average Temperature",
     ylab = "Frequency",
     col = "lightgreen",
     border = "white")

```

Average temperature vs unit_solds.

```{r error=TRUE}

ggplot(df_top3, aes(x = tavg, y = mean(units))) +
  geom_bin2d(bins = 5) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "2D Histogram: Temperature vs Units",
    x = "Average Temperature",
    y = "Units Sold"
  ) +
  theme_minimal()

```

Visualisation of the Month vs Sales

First, aggregate units by month and then Convert month to factor with proper order (1 = Jan, 12 = Dec). and plot the graph.
```{r error=TRUE}

monthly_data <- df_top3 %>%
  group_by(month) %>%
  summarise(units = sum(units, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(month = factor(month, levels = 1:12, labels = month.abb))


```


Monthly top3 item unit sold.

```{r}


# Filter for top 3 items
df_top3_items <- df %>%
  filter(item_nbr %in% c(5, 9, 45))

# Extract month from date
df_top3_items$month <- month(ymd(df_top3_items$date))

# Aggregate units by month and item
monthly_item_data <- df_top3_items %>%
  group_by(month, item_nbr) %>%
  summarise(units = sum(units, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(month = factor(month, levels = 1:12, labels = month.abb),
         item_nbr = factor(item_nbr))

# View the data
print(monthly_item_data)

```

Visualisation of the monthly sales.

```{r}
# Option 4: Line Chart (shows trends clearly)
ggplot(monthly_item_data, aes(x = month, y = units, color = item_nbr, group = item_nbr)) + 
  geom_line(size = 1) +
  geom_point(size = 3) +
  geom_text(aes(label = scales::comma(units)),
            vjust = -1, size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("#378ADD", "#E74C3C", "#2ECC71"),
                     name = "Item Number") +
  labs(
    title    = "Monthly sales trends for top 3 items",
    subtitle = "Items: 5, 9, and 45",
    x        = "Months",
    y        = "Total units sold"
  ) +
  theme_minimal(base_size = 15) +
  theme(
    legend.position = "top"
  )
```

Unit sold by the weather conditions.

```{r}
weather_cols <- c("is_clear","is_rainy","is_snowy","is_foggy","is_stormy","is_weekends", "is_weekdays")

weather_units <- df_top3 %>% pivot_longer(
  cols = all_of(weather_cols),
  names_to = "condition",
  values_to = "present") %>% 
  filter(present == 1) %>%
  group_by(condition) %>% 
             summarise(total_units = sum(units, na.rm = TRUE))

ggplot(weather_units, aes(x = condition, y= total_units)) +
  geom_col(fill = "darkgreen") +
  geom_text(
    aes(label = scales::comma(total_units)),
    vjust = -0.5
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Units Sold by Weather Condition",
    x = "Weather Condition",
    y = "Total Units Sold"
  ) +
  theme_minimal()
             

```

selecting the correlation features 

```{r}
selected_cols <- df_top3 %>% select(
  where(is.numeric)
)# %>% 
  # select(
  #   -c(is_weekends, is_weekdays, is_clear, is_rainy, is_snowy, is_foggy, is_stormy)
  # )

```


```{r}
correlation <- cor(selected_cols, use = "complete.obs", method = "pearson")

# print(round(correlation, 2))
```


```{r}
corrplot(correlation, method = "color", type = "full",
         tl.cex = 0.7, number.cex = 0.6, addCoef.col = "black")
```


## Exploring the Distribution of Sales

Summary statistics are calculated for the "units" variable to understand its 
distribution. Theses includes the minimum, maximum, mean, median, percentage of 
zero sales, and selected percentile values, which help identify skewness and
potential outliers.

```{r error=TRUE}
df_top3 %>%
  summarise(
    min      = min(units, na.rm = TRUE),
    max      = max(units, na.rm = TRUE),
    mean     = round(mean(units, na.rm = TRUE), 2),
    median   = median(units, na.rm = TRUE),
    pct_zero = round(mean(units == 0) * 100, 2),
    p95      = quantile(units, 0.95, na.rm = TRUE),
    p99      = quantile(units, 0.99, na.rm = TRUE),
    p99_5    = quantile(units, 0.995, na.rm = TRUE)
  )

```


```{r}
# write.csv(df_clean, "product_insights.csv", row.names = FALSE)

```


## Data Preparation

Applying a Log Transformation

A logarithmic transformation ("log1p") is applied to the sales variable to reduce
the effects of skewness and make the distribution more suitable for statistical
modelling.

```{r error=TRUE}
df_top3 <- df_top3 %>%
  mutate(units_log = log1p(units))

```



 # Feature Selections
```{r error=TRUE}
features <-  c("tavg","is_clear","is_rainy","is_snowy","is_foggy","is_stormy","is_weekends", "is_weekdays")
```


## Train, Validation and Test the model

Making the data ready for to fit the model. Each product number are set.


One of the top 3 item, item number 5 is filtered 
```{r error=TRUE}
item_data_5 <- df_top3 %>% 
  filter(item_nbr == 5) %>% 
  select(units_log, all_of(features)) %>% na.omit()
```


One of the top 3 item, item number 9 is filtered 
```{r error=TRUE}
item_data_9 <- df_top3 %>% 
  filter(item_nbr == 9) %>%
  select(units_log, all_of(features)) %>% na.omit()
```

One of the top 3 item, item number 5 is filtered 
```{r error=TRUE}
item_data_45 <- df_top3 %>% 
  filter(item_nbr == 45) %>%
  select(units_log, all_of(features)) %>% na.omit()
```

In this section, number of rows for top 3 items model number of rows and different indices are created.

```{r error=TRUE}
set.seed(42)

n5 <- nrow(item_data_5)
n9 <- nrow(item_data_9)
n45 <- nrow(item_data_45)

index_5 <- sample(seq(n5))
index_9 <- sample(seq(n9))
index_45 <- sample(seq(n45))
```


Creating the split for the product no 5
```{r error=TRUE}
# Create Train/ Validation /Test Split
train_5 <- floor(0.8 * n5)
val_5   <- floor(0.9 * n5)

train_data_5 <- item_data_5[index_5[1:train_5], ]
val_data_5   <- item_data_5[index_5[(train_5 + 1):val_5], ]
test_data_5  <- item_data_5[index_5[(val_5 + 1):n5], ]
```

Creating the split for the product no 9
```{r error=TRUE}
# Create Train/ Validation /Test Split
train_9 <- floor(0.8 * n9)
val_9   <- floor(0.9 * n9)

train_data_9 <- item_data_9[index_9[1:train_9], ]
val_data_9   <- item_data_9[index_9[(train_9 + 1):val_9], ]
test_data_9  <- item_data_9[index_9[(val_9 + 1):n9], ]
```

Creating the split for the product no 45
```{r error=TRUE}
# Create Train/ Validation /Test Split
train_45 <- floor(0.8 * n45)
val_45   <- floor(0.9 * n45)

train_data_45 <- item_data_45[index_45[1:train_45], ]
val_data_45   <- item_data_45[index_45[(train_45 + 1):val_45], ]
test_data_45  <- item_data_45[index_45[(val_45 + 1):n45], ]
```



## Fitting the linear regression model for product 5

```{r error=TRUE}
lm_model_5 <- lm(units_log ~ ., data = train_data_5)

```

Validation test and metric to measure for Linear Regression for Product 5.
```{r}
# Validation performance
val_preds_5 <- predict(lm_model_5, newdata = val_data_5)
val_rmse_5 <- sqrt(mean((val_preds_5 - val_data_5$units_log)^2))

```

Testing the model and the metric is used, also train metric for the measurement for Product 5.
```{r}
# Final test performance
test_preds_5 <- predict(lm_model_5, newdata = test_data_5)
test_rmse_5 <- sqrt(mean((test_preds_5 - test_data_5$units_log)^2))
test_mae_5  <- mean(abs(test_preds_5 - test_data_5$units_log))
train_r2_5 <- summary(lm_model_5)$r.squared
```


Displaying the coefficient of product 5.

```{r error=TRUE}
round(coef(lm_model_5), 3)
```

Displaying the error rate for product 5.

```{r error=TRUE}

tibble(
  metric = c("Training R2","Validation RMSE", "Testing RMSE", "Testing MAE"),
  value  = round(c(train_r2_5,val_rmse_5, test_rmse_5, test_mae_5), 3)
)

```



Fitting the Linear Regression Model for Product 9
```{r error=TRUE}
lm_model_9 <- lm(units_log ~ ., data = train_data_9)

```

Validation test and metric to measure for Linear Regression for Product 9.
```{r}
# Validation performance
val_preds_9 <- predict(lm_model_9, newdata = val_data_9)
val_rmse_9 <- sqrt(mean((val_preds_9 - val_data_9$units_log)^2))
```

Testing the model and the metric is used, also train metric for the measurement for Product 9.
```{r}
# Final test performance
test_preds_9 <- predict(lm_model_9, newdata = test_data_9)
test_rmse_9 <- sqrt(mean((test_preds_9 - test_data_9$units_log)^2))
test_mae_9  <- mean(abs(test_preds_9 - test_data_9$units_log))
train_r2_9 <- summary(lm_model_9)$r.squared
```


Displaying the coefficient of product 9.

```{r error=TRUE}
round(coef(lm_model_9), 3)
```

Displaying the error rate.

```{r error=TRUE}

tibble(
  metric = c("Training R2","Validation RMSE", "Testing RMSE", "Testing MAE"),
  value  = round(c(train_r2_9,val_rmse_9, test_rmse_9, test_mae_9), 3)
)

```


Fitting the Linear Regression Model for Product 45.
```{r error=TRUE}
lm_model_45 <- lm(units_log ~ ., data = train_data_45)

```

Validation test and metric to measure for Linear Regression for Product 45.
```{r}
# Validation performance
val_preds_45 <- predict(lm_model_45, newdata = val_data_45)
val_rmse_45 <- sqrt(mean((val_preds_45 - val_data_45$units_log)^2))
```

Testing the model and the metric is used, also train metric for the measurement for Product 45.
```{r}
# Final test performance
test_preds_45 <- predict(lm_model_45, newdata = test_data_45)
test_rmse_45 <- sqrt(mean((test_preds_45 - test_data_45$units_log)^2))
test_mae_45  <- mean(abs(test_preds_45 - test_data_45$units_log))
train_r2_45 <- summary(lm_model_45)$r.squared
```


Displaying the coefficient of product 45.

```{r error=TRUE}
round(coef(lm_model_45), 3)
```

Displaying the error rate.

```{r error=TRUE}

tibble(
  metric = c("Training R2","Validation RMSE", "Testing RMSE", "Testing MAE"),
  value  = round(c(train_r2_45,val_rmse_45, test_rmse_45, test_mae_45), 3)
)

```


# Decision Tree Models

## Decision Tree 5

```{r error=TRUE}
dt_model_5 <- rpart(
  units_log ~ .,
  data = train_data_5,
  method = "anova",   
  control = rpart.control(cp = 0.001, 
                          minsplit = 15
                          )
  )
```

The decision trees regression training and validation and measuring training r2 and validation rmse to measure the error for model 5.
```{r error=TRUE}

dt_train_preds_5 <- predict(dt_model_5, newdata = train_data_5)
dt_train_r2_5 <- 1 - sum((train_data_5$units_log - dt_train_preds_5)^2) /
                  sum((train_data_5$units_log - mean(train_data_5$units_log))^2)

dt_val_preds_5  <- predict(dt_model_5, newdata = val_data_5)
dt_val_rmse_5   <- sqrt(mean((dt_val_preds_5 - val_data_5$units_log)^2))

```

Running the prediction, test root mean square error(rmse) and mean absolute error(mae) to measure the testing result for model 5.
```{r error=TRUE}
dt_test_preds_5 <- predict(dt_model_5, newdata = test_data_5)
dt_test_rmse_5  <- sqrt(mean((dt_test_preds_5 - test_data_5$units_log)^2))
dt_test_mae_5   <- mean(abs(dt_test_preds_5 - test_data_5$units_log))

```

Metrics for the decision tree regression model 5.
```{r}
tibble(
  metric = c("Training R2","Validation RMSE", "Testing RMSE", "Testing MAE"),
  value  = round(c(dt_train_r2_5,dt_val_rmse_5, dt_test_rmse_5, dt_test_mae_5), 3)
)
```

#Pruning the Tree

```{r}

# Visualize pruned tree
rpart.plot(dt_model_5, main = "Visualization Decision Tree for Model 5 Regression")

```


```{r}
round(dt_model_5$variable.importance, 3)
```


Visualization of the importance on the Model 5.
```{r}
barplot(dt_model_5$variable.importance,
        main = "Feature Importance Model 5 (Decision Tree)",
        ylab = " Feature Numbers",
        las = 2,
        col = "steelblue")

```



## Decision Tree 9

```{r error=TRUE}
dt_model_9 <- rpart(
  units_log ~ .,
  data = train_data_9,
  method = "anova",              # regression tree (units is numeric)
  control = rpart.control(cp = 0.001, 
                          minsplit = 20
                          )
  )
```

The decision trees regression training and validation and measuring training r2 and validation rmse to measure the error for model 9.
```{r error=TRUE}

dt_train_preds_9 <- predict(dt_model_9, newdata = train_data_9)
dt_train_r2_9 <- 1 - sum((train_data_5$units_log - dt_train_preds_9)^2) /
                  sum((train_data_9$units_log - mean(train_data_9$units_log))^2)

dt_val_preds_9  <- predict(dt_model_9, newdata = val_data_9)
dt_val_rmse_9   <- sqrt(mean((dt_val_preds_9 - val_data_9$units_log)^2))

```

Running the prediction, test root mean square error(rmse) and mean absolute error(mae) to measure the testing result for model 9.
```{r error=TRUE}
dt_test_preds_9 <- predict(dt_model_9, newdata = test_data_9)
dt_test_rmse_9  <- sqrt(mean((dt_test_preds_9 - test_data_9$units_log)^2))
dt_test_mae_9   <- mean(abs(dt_test_preds_9 - test_data_9$units_log))

```

Metrics for the Decision Tree Regression model for model 9.
```{r}
tibble(
  metric = c("Training R2","Validation RMSE", "Testing RMSE", "Testing MAE"),
  value  = round(c(dt_train_r2_9,dt_val_rmse_9, dt_test_rmse_9, dt_test_mae_9), 3)
)
```

Pruning the Tree and Visualize pruned tree

```{r}
rpart.plot(dt_model_9, main = "Visualisation Decision Tree for Model 9 Regression")

```


```{r}
round(dt_model_9$variable.importance, 3)
```


Visualization of the importance on the Model 5.
```{r}
barplot(dt_model_9$variable.importance,
        main = "Feature Importance  for Model 9 (Decision Tree)",
        ylab = " Features Number",
        las = 2,
        col = "steelblue")

```



## Decision Tree 45

```{r error=TRUE}
dt_model_45 <- rpart(
  units_log ~ .,
  data = train_data_45,
  method = "anova",              # regression tree (units is numeric)
  control = rpart.control(cp = 0.001, 
                          minsplit = 15
                          )
  )
```


The decision trees regression training and validation and measuring training r2 and validation rmse to measure the error for model 45.
```{r error=TRUE}

dt_train_preds_45 <- predict(dt_model_45, newdata = train_data_45)
dt_train_r2_45 <- 1 - sum((train_data_45$units_log - dt_train_preds_45)^2) /
                  sum((train_data_45$units_log - mean(train_data_45$units_log))^2)

dt_val_preds_45  <- predict(dt_model_45, newdata = val_data_45)
dt_val_rmse_45   <- sqrt(mean((dt_val_preds_45 - val_data_45$units_log)^2))

```


Running the prediction, test root mean square error(rmse) and mean absolute error(mae) to measure the testing result for model 45.
```{r error=TRUE}
dt_test_preds_45 <- predict(dt_model_45, newdata = test_data_45)
dt_test_rmse_45  <- sqrt(mean((dt_test_preds_45 - test_data_45$units_log)^2))
dt_test_mae_45   <- mean(abs(dt_test_preds_45 - test_data_45$units_log))

```

Metrics for the Decision Tree Regression model for model 9.
```{r}
tibble(
  metric = c("Training R2","Validation RMSE", "Testing RMSE", "Testing MAE"),
  value  = round(c(dt_train_r2_45,dt_val_rmse_45, dt_test_rmse_45, dt_test_mae_45), 3)
)
```


Pruning the Tree and Visualize pruned tree
```{r}
rpart.plot(dt_model_45, main = "Visualisation Decision Tree Regression for  Model 45")
```

looking for the importance of the features used for the prediction model 45.
```{r}
round(dt_model_45$variable.importance, 3)
```


Visualization of the importance on the Model 45.
```{r error=TRUE}
barplot(dt_model_45$variable.importance,
        main = "Feature Importance for Model 45 (Decision Tree)",
        ylab = "Importance Number",
        las = 2,
        col = "steelblue")

```





 What's Missing Entirely
 
- Written interpretations for each product

- Business recommendations section

- Proper handling of "T" values in weather data

- Justification for outlier approach (especially zero-inflation)

- Proper visuals with clear interpretation

- Identification of top 3 products via SQL


 Recommendations to Complete the Assignment
 
- Add SQL query to find top 3 products (Section 4.1)

- Fix data cleaning for "T" values and outliers

- Add written analysis for each of the 3 products

- Create proper business recommendations

- Add better visualizations with proper titles/labels

- Make code reproducible using relative paths

- Fix the Product 45 model data reference



















