---
title: "R Notebook"
output: html_notebook
---

```{r}
install.packages("tidyr")
install.packages("tidyverse")
```

```{r message=FALSE, warning=FALSE, error=TRUE}
library(DBI)
library(RSQLite)
library(dplyr)      
library(lubridate)   
library(ggplot2)    
library(yardstick)  
library(dplyr)
library(readr)
library(tidyr)
library(tidyverse)
```


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.


```{r error=TRUE}
con <- dbConnect(RSQLite::SQLite(),
                 "/Users/AshishLama/Desktop/project1_raw_data.db") 

#dbListFields(con, "station")



data = dbGetQuery(con, "PRAGMA table_info(weather)")
```

```{r error=TRUE}
df <- dbGetQuery(con, "
SELECT t.date, t.store_nbr, t.item_nbr, t.units,
       w.tavg, w.preciptotal, w.codesum, w.dewpoint, w.wetbulb,  w.heat, w.cool
FROM train t
JOIN key k ON t.store_nbr = k.store_nbr
JOIN weather w ON k.station_nbr = w.station_nbr
               AND t.date = w.date
")
```


view the Dataframe

```{r error= TRUE}
df
View(df)
```

closing the connection

```{r error=TRUE}
dbDisconnect(con)
```

```{r error=TRUE}
summary(df)
```





```{r}
names(df)
```

```{r}
sum(is.na(df$tavg))  # Count NA values
sum(df$tavg == " ") 
```


```{r}
df_clean <- df %>%
  mutate(
    # Convert empty strings to NA
    tavg = na_if(tavg, 0),
    # Calculate median (store it)
    median_tavg = median(tavg, na.rm = TRUE),
    # Replace NAs with median
    tavg = ifelse(is.na(tavg), median_tavg, tavg)
  ) %>%
  select(-median_tavg)  # Remove helper column

# Check results
summary(df_clean$tavg)
```

```{r}
df_clean
```

```{r}
Sys.Date()
df_clean$date <- as.Date(df_clean$date, format = "%Y-%m-%d")

```


```{r}
df_clean$day <- weekdays(df_clean$date)
df_clean$month = as.integer(format(df_clean$date, "%m"))
df_clean$year  = as.integer(format(df_clean$date, "%Y"))
```





```{r}
df_clean
```

```{r}
df_clean <- df_clean %>%
  mutate(codesum = ifelse(codesum == "" | codesum == " ", "Moderate", codesum))

# Check the results
#table(df_clean$codesum)
```

```{r}
df_clean
```

```{r}
print(nrow(df_clean))
```

```{r}
sapply(df_clean, class)
```




```{r}
june <- df_clean %>% filter(month == 6)
june
```


```{r}
# Summarise total units by month
monthly_sales <- df_clean %>%
  group_by(month) %>%
  summarise(total_units = sum(units, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    month_name = month.abb[month],
  )

# Print the comparison
monthly_sales
```

```{r}
library(ggplot2)

ggplot(monthly_sales, aes(x = reorder(month_name, month), 
                           y = total_units)) + geom_col(fill = "#378ADD") +
  geom_text(aes(label = scales::comma(total_units)), 
            vjust = -0.5, size = 3.2)  +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Total units sold by month",
    x        = NULL,
    y        = "Total units sold",
    fill     = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    legend.position    = "top"
  )
```


```{r}
df_clean <- df_clean %>%
  mutate(
    is_clear  = if_else(str_detect(codesum, "Moderate"), 1L, 0L),
    is_rainy  = if_else(str_detect(codesum, "RA|DZ"), 1L, 0L),
    is_snowy  = if_else(str_detect(codesum, "SN"), 1L, 0L),
    is_foggy  = if_else(str_detect(codesum, "FG|BR"), 1L, 0L),
    is_stormy = if_else(str_detect(codesum, "TS"), 1L, 0L)
  )
```

```{r}
df_clean
```


```{r}

```

```{r}

```








