---
title: "R Notebook"
output: html_notebook
---

```{r}
install.packages("tidyr")
install.packages("tidyverse")
```

```{r message=FALSE, warning=FALSE, error=TRUE}
library(DBI)
library(RSQLite)
library(dplyr)      
library(lubridate)   
library(ggplot2)    
library(yardstick)  
library(dplyr)
library(readr)
library(tidyr)
library(tidyverse)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

```{r error=TRUE}
con <- dbConnect(RSQLite::SQLite(),
                 "/Users/AshishLama/Desktop/project1_raw_data.db") 

#dbListFields(con, "station")



data = dbGetQuery(con, "PRAGMA table_info(weather)")
```

```{r error=TRUE}
df <- dbGetQuery(con, "
SELECT t.date, t.store_nbr, t.item_nbr, t.units,
       w.tavg, w.preciptotal, w.codesum, w.dewpoint, w.wetbulb,  w.heat, w.cool
FROM train t
JOIN key k ON t.store_nbr = k.store_nbr
JOIN weather w ON k.station_nbr = w.station_nbr
               AND t.date = w.date
")
```

view the Dataframe

```{r error= TRUE}
df
View(df)
```

closing the connection

```{r error=TRUE}
dbDisconnect(con)
```

```{r error=TRUE}
summary(df)
```

```{r}
names(df)
```

```{r}
sum(is.na(df$tavg))  # Count NA values
sum(df$tavg == " ") 
```

```{r}
df_clean <- df %>%
  mutate(
    # Convert empty strings to NA
    tavg = na_if(tavg, 0),
    # Calculate median (store it)
    median_tavg = median(tavg, na.rm = TRUE),
    # Replace NAs with median
    tavg = ifelse(is.na(tavg), median_tavg, tavg)
  ) %>%
  select(-median_tavg)  # Remove helper column

# Check results
summary(df_clean$tavg)
```

```{r}
df_clean
```

```{r}
Sys.Date()
df_clean$date <- as.Date(df_clean$date, format = "%Y-%m-%d")

```

```{r}
df_clean$day <- weekdays(df_clean$date)
df_clean$month = as.integer(format(df_clean$date, "%m"))
```

```{r}
df_clean
```

```{r}
df_clean <- df_clean %>%
  mutate(codesum = ifelse(codesum == "" | codesum == " ", "Moderate", codesum))

# Check the results
#table(df_clean$codesum)
```

```{r}
df_clean
```

```{r}
#write.csv(df_clean, "data.csv", row.names = FALSE)
```

```{r}
print(nrow(df_clean))
```

```{r}
sapply(df_clean, class)
```

```{r}
june <- df_clean %>% filter(month == 6)
june
```

```{r}
# Summarise total units by month
monthly_sales <- df_clean %>%
  group_by(month) %>%
  summarise(total_units = sum(units, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    month_name = month.abb[month],
  )

# Print the comparison
monthly_sales
```

```{r}
library(ggplot2)

ggplot(monthly_sales, aes(x = reorder(month_name, month), 
                           y = total_units)) + geom_col(fill = "#378ADD") +
  geom_text(aes(label = scales::comma(total_units)), 
            vjust = -0.5, size = 3.2)  +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Total units sold by month",
    x        = NULL,
    y        = "Total units sold",
    fill     = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    legend.position    = "top"
  )
```

```{r}
df_clean <- df_clean %>%
  mutate(
    is_clear  = if_else(str_detect(codesum, "Moderate"), 1L, 0L),
    is_rainy  = if_else(str_detect(codesum, "RA|DZ"), 1L, 0L),
    is_snowy  = if_else(str_detect(codesum, "SN"), 1L, 0L),
    is_foggy  = if_else(str_detect(codesum, "FG|BR"), 1L, 0L),
    is_stormy = if_else(str_detect(codesum, "TS"), 1L, 0L)
  )
```

```{r}
df_clean
```

```{r}
# Step 1 — Understand the distribution first
df_clean %>%
  summarise(
    min      = min(units, na.rm = TRUE),
    max      = max(units, na.rm = TRUE),
    mean     = round(mean(units, na.rm = TRUE), 3),
    median   = median(units, na.rm = TRUE),
    pct_zero = round(mean(units == 0) * 100, 1),
    p95      = quantile(units, 0.95, na.rm = TRUE),
    p99      = quantile(units, 0.99, na.rm = TRUE),
    p99_5    = quantile(units, 0.995, na.rm = TRUE)
  )
```

```{r}
nonzero_units <- df_clean$units[df_clean$units > 0]

nonzero_summary <- tibble(
  n_nonzero = length(nonzero_units),
  min       = min(nonzero_units),
  p50       = quantile(nonzero_units, 0.50),
  p90       = quantile(nonzero_units, 0.90),
  p95       = quantile(nonzero_units, 0.95),
  p99       = quantile(nonzero_units, 0.99),
  p99_5     = quantile(nonzero_units, 0.995),
  p99_9     = quantile(nonzero_units, 0.999),
  max       = max(nonzero_units)
)

nonzero_summary
```

```{r}
upper_cap <- quantile(nonzero_units, 0.995)
cat("Cap threshold (p99.5 of non-zero sales):", upper_cap, "\n")

df_clean <- df_clean %>%
  mutate(
    is_outlier   = if_else(units > upper_cap, 1, 0),
    units_capped = pmin(units, upper_cap)
  )

# How many rows are affected?
df_clean %>%
  summarise(
    n_outliers = sum(is_outlier),
    pct_outliers = round(mean(is_outlier) * 100, 3)
  )
```

```{r}
df_clean <- df_clean %>%
  mutate(units_log = log1p(units))

range(df_clean$units_log)
```

```{r}
install.packages("patchwork")
```

```{r}
library(ggplot2)
library(patchwork)

# Bin the units for a clearer view (raw histogram is unreadable at 97.4% zeros)
df_plot <- df_clean %>%
  mutate(
    units_bin = case_when(
      units == 0    ~ "0",
      units <= 37   ~ "1-37 (≤p99)",
      units <= 66   ~ "38-66 (p99-p99.5)",
      TRUE          ~ ">66 (outliers)"
    ),
    units_bin = factor(units_bin, levels = c("0","1-37 (≤p99)","38-66 (p99-p99.5)",">66 (outliers)"))
  )

# Panel 1 — original distribution, log scale to make small bars visible
p1 <- df_plot %>%
  count(units_bin) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = units_bin, y = pct, fill = units_bin)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct, 2), "%")), vjust = -0.3, size = 3.2) +
  scale_y_log10() +
  scale_fill_manual(values = c("#85B7EB","#378ADD","#185FA5","#D85A30")) +
  labs(title = "Original units distribution (log scale)", x = NULL, y = "% of rows (log scale)") +
  theme_minimal()

# Panel 2 — after capping at p99.5
p2 <- ggplot(df_clean, aes(x = units_capped)) +
  geom_histogram(binwidth = 2, fill = "#185FA5") +
  scale_y_log10() +
  labs(title = paste0("After capping at p99.5 (", upper_cap, ")"), 
       x = "units_capped", y = "count (log scale)") +
  theme_minimal()

# Panel 3 — log1p transform
p3 <- ggplot(df_clean, aes(x = units_log)) +
  geom_histogram(binwidth = 0.3, fill = "#0C447C") +
  labs(title = "log1p(units) transform", x = "log1p(units)", y = "count") +
  theme_minimal()

p1 / p2 / p3
```

```{r}
df_clean
```

```{r}
df_clean$day <- as.integer(format(df_clean$date, "%u"))

df_clean

```

```{r}
# Setup and feature selection
library(dplyr)
library(rpart)
library(randomForest)
library(xgboost)
library(Metrics)
library(data.table)
library(mlbench)
library(deepnet)
library(caTools)

set.seed(123)

# Select relevant features + target
model_data <- df_clean %>%
  select(units_capped,                       # target (capped at p99.5)
         store_nbr, item_nbr, 
         tavg, preciptotal,
         is_clear, is_rainy, is_snowy, is_foggy, is_stormy, month) #%>% na.omit()

# Train/test split (80/20)
train_idx <- sample(seq_len(nrow(model_data)), size = 0.8 * nrow(model_data))
train <- model_data[train_idx, ]
test  <- model_data[-train_idx, ]
```

```{r}
# Linear Regression
lm_model <- lm(units_capped ~ store_nbr + item_nbr + #tavg + preciptotal  +
                 #is_clear + is_rainy + is_snowy + is_foggy + is_stormy +
                 month + day,
               data = train)

summary(lm_model)

lm_pred <- predict(lm_model, newdata = test)
```

```{r}
tree_model <- rpart(units_capped ~ store_nbr + item_nbr + tavg + preciptotal +
                      is_clear + is_rainy + is_snowy + is_foggy + is_stormy +
                       month,
                    data = train, method = "anova")

# Visualize the tree
library(rpart.plot)
rpart.plot(tree_model, main = "Decision Tree for Units Sold")

tree_pred <- predict(tree_model, newdata = test)
```


```{r}
#Logistic model

logistic_model <- glm(units_capped ~ store_nbr + item_nbr + tavg + preciptotal +
                        is_clear + is_rainy + is_snowy + is_foggy + is_stromy +
                        month)



```



```{r}

```


```{r}

```












```{r}
# rf_model <- randomForest(units_capped ~ store_nbr + item_nbr + tavg + preciptotal +
#                            is_clear + is_rainy + is_snowy + is_foggy + is_stormy +
#                             month,
#                          data = train, ntree = 100, importance = TRUE)
# 
# rf_pred <- predict(rf_model, newdata = test)
# 
# # Feature importance
# importance(rf_model)
varImpPlot(rf_model)
```

```{r}
# # XGBoost requires a matrix, not a formula
# feature_cols <- c("store_nbr","item_nbr","tavg","preciptotal",
#                    "is_clear","is_rainy","is_snowy","is_foggy","is_stormy","month")
# 
# train_matrix <- as.matrix(train[, feature_cols])
# test_matrix  <- as.matrix(test[, feature_cols])
# 
# dtrain <- xgb.DMatrix(data = train_matrix, label = train$units_capped)
# dtest  <- xgb.DMatrix(data = test_matrix, label = test$units_capped)
# 
# xgb_model <- xgboost(
#   x       = dtrain,
#   nrounds    = 100,
#   max_depth  = 6,
#   learning_rate = 0.01,
#   objective  = "reg:squarederror"
#   #verbose    = 0
# )
# 
# xgb_pred <- predict(xgb_model, dtest)
# 
# # Feature importance
# xgb.importance(model = xgb_model)
```

```{r}
results <- data.frame(
  Model = c("Linear Regression", "Decision Tree"), #"Random Forest", "XGBoost"),
  RMSE  = c(rmse(test$units_capped, lm_pred),
            rmse(test$units_capped, tree_pred)),
            #rmse(test$units_capped, rf_pred),
            #rmse(test$units_capped, xgb_pred)),
  MAE   = c(mae(test$units_capped, lm_pred),
            mae(test$units_capped, tree_pred)),
            #mae(test$units_capped, rf_pred),
           # mae(test$units_capped, xgb_pred)),
  R2    = c(cor(test$units_capped, lm_pred)^2,
            cor(test$units_capped, tree_pred)^2)
            #cor(test$units_capped, rf_pred)^2,
            #cor(test$units_capped, xgb_pred)^2)
)

results
```

```{r}

```
